

# This is a header

## This is a subheader

Bold like **this**

Italics like *this*

Links like [this](http://www.google.com)

Inline code like this `myVar=1`

Unordered lists like this:

- item 1
- item 2
- item 3

Ordered lists like this:

1. item 1
2. item 2
3. item 3

Inline math like this \(x^2\)

or like this $x^2$

Block math like this:

$$
f(x) = \frac{1}{(\sigma\sqrt{2 \pi})} e^{-(\frac{(x - \mu)^2}{2 \sigma^2})}
$$

or like this:

\[
f(x) = \frac{1}{(\sigma\sqrt{2 \pi})} e^{-(\frac{(x - \mu)^2}{2 \sigma^2})}
\]

Inline R output like this: `r rnorm(1)`.

Block R code like this:

```{r}
f <- function(n) rnorm(n)
f(10)
```

Plots like this:

```{r}
plot(1:10)
```

Plain old html:
<a href="http://google.com">google</a>

Syntax-highlighted code blocks (many languages available) like this:

<pre><code class="r">f <- function(x) {
   cat("hi")
   x <- c(1:10)
   rnorm(n)
}</code></pre>

A list of displays from your VDB:

```{r echo=FALSE, results="asis"}
nbDisplayList(c("plot1", "plot2", "plot3"))
```

A single display from your VDB:

```{r echo=FALSE, results="asis"}
nbDisplayList("plot4")
```

More code for fun...

A Cross Product Formula:

\[
  \mathbf{V}_1 \times \mathbf{V}_2 =
   \begin{vmatrix}
    \mathbf{i} & \mathbf{j} & \mathbf{k} \\
    \frac{\partial X}{\partial u} & \frac{\partial Y}{\partial u} & 0 \\
    \frac{\partial X}{\partial v} & \frac{\partial Y}{\partial v} & 0 \\
   \end{vmatrix}
\]

The probability of getting \(k\) heads when flipping \(n\) coins is:

\[ P(E) = {n \choose k} p^k (1-p)^{ n-k} \]


